Flat-histogram method comparison on 2D Ising model

### 2020-05-04 ising/ising-body.tex
added coupling constant J and made beta/J for figure. Changed acknowledgement.

\section{Ising Model}
The 2D Ising spin-lattice system is widely used as a testbed when
benchmarking or comparing Monte Carlo
methods~\cite{ferdinand1969bounded, wang1999transition, trebst2004optimizing, barash2019estimating}. The 2nd order
phase transition behavior and the ability to directly calculate the
exact solution for finite lattices~\cite{beale1996exact, haggkvist2004computation} make the
system sufficiently interesting for such theoretical comparisons. It is
also important to note that direct comparison of the other methods can
be made with WL as its original implementation was done on this
system~\cite{wang2001determining,wang2001efficient}.
We test the convergence of several flat-histogram methods
on the periodic 2D square lattice ferromagnetic Ising model with nearest
neighbor interactions~\cite{landau2004new}.
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j - h \sum_i s_i
\end{align}
The $N\times N$ spin system can take on values of $\sigma_i = \pm 1$
for up or down spins respectively. In the absence of a magnetic field ($h =
0$), We can write the Hamiltonian as follows~\cite{onsager1944crystal,
kaufman1949crystal}:
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j
\end{align}
where the sum is over nearest neighbor spin sites. Beale showed that for finite
lattices the Density of States could directly be calculated from the partition
function~\cite{beale1996exact}
\begin{align}
Z = \sum_E g(E) e^{-{\beta E}}
\end{align}
% where $g(E)$ is the multiplicity of the system which is proportional to
% $D(E)$. We can estimate the average deviation from the exact solution
% using the relation~\cite{schneider2017convergence, shakirov2018convergence,
% barash2017control}:
% \begin{align}
% \braket{\epsilon(t)}_E = \frac{1}{N_E - 1} \sum_E \left( \frac{S(E,t) - S_{\text{Beale}(E)}}{S_{\text{Beale}(E)}} \right)
% \end{align}
where $g(E)$ is the multiplicity of the system which is proportional to
$D(E)$. We can compute the maximum deviation in the canonical specific heat capacity $c_V$ from the exact solution~\cite{schneider2017convergence, shakirov2018convergence,
barash2017control,barash2017gpu}:
\begin{align}
\epsilon \equiv \max_T \left|\left(c_V(T)_{\text{MC}} - c_V(T)_{\text{Beale}}\right)\right|
\end{align}
% As per the implementation by Wang and
% Landau~\cite{wang2001determining}, the total number of available energy
% states are $N-1$.
Computing the specific heat capacity $c_V$ presents a difficult challenge for
any Monte Carlo method due to fluctuations in the derivative of the
internal energy.  Methods that accurately compute $c_V$ also by
extension accurately compute the internal energy.

@@

We thank Klas Markstr\"{o}m for discussions regarding the exact computation of the partition function for the 2D Ising model. We also wish to thank Johannes Zierenberg for helpful discussions regarding WL production run simulations and
Martin Weigel for discussions regarding entropic population annealing.

>>>>

\section{Ising Model}
The 2D Ising spin-lattice system is widely used as a testbed when
benchmarking or comparing Monte Carlo
methods~\cite{ferdinand1969bounded, wang1999transition, trebst2004optimizing, barash2019estimating}. The 2nd order
phase transition behavior and the ability to directly calculate the
exact solution for finite lattices~\cite{beale1996exact, haggkvist2004computation} make the
system sufficiently interesting for such theoretical comparisons. It is
also important to note that direct comparison of the other methods can
be made with WL as its original implementation was done on this
system~\cite{wang2001determining,wang2001efficient}.
We test the convergence of several flat-histogram methods
on the periodic 2D square lattice ferromagnetic Ising model with identical
nearest neighbor interactions~\cite{landau2004new} ($J_{ij} = J$).
\begin{align}
\mathcal{H} = -J \sum_{\braket{i,j}} \sigma_i \sigma_j - h \sum_i s_i
\end{align}
The $N\times N$ spin system can take on values of $\sigma_i = \pm 1$
for up or down spins respectively. In the absence of a magnetic field ($h =
0$), We can write the Hamiltonian as follows~\cite{onsager1944crystal,
kaufman1949crystal}:
\begin{align}
\mathcal{H} = - J\sum_{\braket{i,j}} \sigma_i \sigma_j
\end{align}
where the sum is over nearest neighbor spin sites. Beale showed that for finite
lattices the Density of States could directly be calculated from the partition
function~\cite{beale1996exact}
\begin{align}
Z = \sum_E g(E) e^{-{\beta E}}
\end{align}
% where $g(E)$ is the multiplicity of the system which is proportional to
% $D(E)$. We can estimate the average deviation from the exact solution
% using the relation~\cite{schneider2017convergence, shakirov2018convergence,
% barash2017control}:
% \begin{align}
% \braket{\epsilon(t)}_E = \frac{1}{N_E - 1} \sum_E \left( \frac{S(E,t) - S_{\text{Beale}(E)}}{S_{\text{Beale}(E)}} \right)
% \end{align}
where $g(E)$ is the multiplicity of the system which is proportional to
$D(E)$. We can compute the maximum deviation in the canonical specific heat capacity $c_V$ from the exact solution~\cite{schneider2017convergence, shakirov2018convergence,
barash2017control,barash2017gpu}:
\begin{align}
\epsilon \equiv \max_T \left|\left(c_V(T)_{\text{MC}} - c_V(T)_{\text{Beale}}\right)\right|
\end{align}
% As per the implementation by Wang and
% Landau~\cite{wang2001determining}, the total number of available energy
% states are $N-1$.
Computing the specific heat capacity $c_V$ presents a difficult challenge for
any Monte Carlo method due to fluctuations in the derivative of the
internal energy.  Methods that accurately compute $c_V$ also by
extension accurately compute the internal energy.

@@

We wish to thank Johannes Zierenberg for helpful discussions regarding WL followed by a production run and for insights into properties of a production run at different stages of convergence.
